---
title             : "Social Media Use in Youth and Related Mental Health and Behavioral Outcomes"
shorttitle        : "Youth Social Media Use"

author: 
  - name          : "Natalie Cunningham"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "75 E River Pkwy, Minneapolis, MN 55455"
    email         : "cunni734@umn.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "University of Minnesota - Twin Cities"
  - id            : "2"
    institution   : "Konstanz Business School"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  One or two sentences to put the results into a more **general context**.
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library(papaja)
library(groundhog)
groundhog.library(dplyr, "2025-04-02")
groundhog.library(psych, "2025-04-02")
groundhog.library(labelled, "2025-04-02")
groundhog.library(ggplot2, "2025-04-02")
groundhog.library(apaTables, "2025-04-02")
r_refs("r-references.bib")
```

```{r data setup, include=FALSE}
dat <- read.csv("../Data/2025-11-13_youth-media-use_simulated-data.csv")
names(dat)
```

## Procedure

The Monitoring the Future Main Study is a survey that is presented annually to eighth, tenth, and twelfth graders to observe how answers may change over time. The survey results are split into two data sets, one for eighth and tenth grade participants and one for twelfth grade participants. For the sake of this study, items were pulled from the 2023 8th- and 10th grade survey codebook to create scales of social media use, impulsivity, depression, and anxiety.

# Results


# Discussion


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
