---
title             : "Social Media Use in Youth and Related Mental Health and Behavioral Outcomes"
shorttitle        : "Youth Social Media Use"

author: 
  - name          : "Natalie Cunningham"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "75 E River Pkwy, Minneapolis, MN 55455"
    email         : "cunni734@umn.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "University of Minnesota - Twin Cities"

abstract: |
  Social media is becoming a new constant in the lives of high school youth, inducing worry among the public regarding the potential behavioral and mental health outcomes that social media use may have on adolescents. More research is required to fully understand the impact that social media use has on the adolescent brain and behavior, and should continue to be conducted as technology advances. This study looks at the relationship between social media and impulsivity, depression, and anxiety in high school youth through simulated data created by means of the 2023 Monitoring the Future continuing study conducted on eighth and tenth grade participants by the University of Michigan. Through the simulated data, it was found that there were no statistically significant relationships between social media use and the measured mental health and behavioral outcomes, nor was social media use seen as a predictor of these outcomes. However, with raw data, different results may be observed in these regards, and further analysis should be done on the raw data to assess these relationships.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "social media use, youth, mental health outcomes, behavioral outcomes"
wordcount         : "1361"

bibliography      : ["r-references.bib", "references.bib"]

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
keep_tex          : yes

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("dplyr")
library("psych")
library("labelled")
library("ggplot2")
library("apaTables")
library("tidyr")
library("cowplot")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(1980)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

```{r data setup, include=FALSE}
dat <- read.csv("../Data/2025-11-13_youth-media-use_simulated-data.csv")
names(dat)
```

# Introduction
Technology is rapidly advancing and becoming more accessible to current adolescents and those of future generations. With this accessibility, the doors to social media access and use are wide open to youth. Simultaneously, however, mental health and behavioral issues among youth are being more widely addressed. Many claims have been made that social media, and media use in general, have negative effects on children's well-being, and new trends related to parenting methods with regard to technology access are widespread [@turner2020]. These technologies and media platforms will continue to evolve, and, therefore, it is becoming increasingly necessary to understand the psychological impacts of these growing entities on youth. For the sake of this notion, I analyzed how social media use parallels adolescent levels of mental health and behavioral outcomes, specifically impulsivity, depression, and anxiety. The focus of this study is to use reproducible workflows to examine these relationships among a population of high-school-aged youth.
Three hypotheses were created for how I expected social media use to impact impulsivity, depression, and anxiety in the surveyed participants. All three hypotheses infer that social media use will be positively correlated with each outcome, respectively, so higher use of social media will correspond with higher levels of impulsivity, depression, and anxiety.
To assess these, analyses were done on a simulated data set compiled referencing the data codebook from the 2023 Monitoring the Future Main Study, conducted on eighth and tenth-grade participants by the University of Michigan [@monitori]. The hypotheses were analyzed using the simulated data to outline the procedure and method by which these research questions could be assessed in a reproducible manner. This project also provides the foundation structure for future analysis of the hypotheses without the confines of simulated data and instead with the provided raw data from which the simulated data set was formulated.

# Methods
The current study was not preregistered, and raw data were not collected or analyzed. Simulated data was created based on raw data from the 2023 study on eighth and tenth graders from the “Monitoring the Future” continuing study. Simulated data and code are available at [this hyperlink to the GitHub repository](https://github.com/cunningpsych/Social-Media-Use_in_Youth).

## Participants
```{r demographics, include=FALSE}
# Using prettyNum() to format the number of participants to read nicer in the output
num_dat <- nrow(dat)
formatted_num <- prettyNum(num_dat, big.mark = ",", scientific = FALSE)

# recoding grade factors from (1,2) to (eighth grade, tenth grade) and creating objects for each grade
dat <- mutate(dat, grade_sim = factor(grade_sim, levels = c(1,2),
                                      labels = c("Eighth Grade", "Tenth Grade")))

summarytools::freq(dat$grade_sim)

grade_n <- ""
grade_percent <- ""
eighth_n <- sum(dat$grade_sim == "Eighth Grade")
eighth_percent <- round(((eighth_n/nrow(dat))*100), digits = 2)
tenth_n <- sum(dat$grade_sim == "Tenth Grade")
tenth_percent <- round(((tenth_n/nrow(dat))*100), digits = 2)


# recoding residence factors from (1,2,3) to (farm, country, citytown), creating objects
dat <- mutate(dat, residence_sim = factor(residence_sim, levels = c(1,2,3),
                                          labels = c("Farm", "Country", "City")))
summarytools::freq(dat$residence_sim)

residence_n <- ""
residence_percent <- ""
farm_n <- nrow(filter(dat, residence_sim == "Farm"))
country_n <- nrow(filter(dat, residence_sim == "Country"))
city_n <- nrow(filter(dat, residence_sim == "City"))
farm_percent <- round(((nrow(filter(dat, residence_sim == "Farm")))/(nrow(dat))*100), digits = 2)
country_percent <- round(((nrow(filter(dat, residence_sim == "Country")))/(nrow(dat))*100), digits = 2)
city_percent <- round(((nrow(filter(dat, residence_sim == "City")))/(nrow(dat))*100), digits = 2)


# recoding sex factors from (1,2,3,4) to (male, female, other, prefer not to answer), creating objects
dat <- mutate(dat, sex_sim = factor(sex_sim, levels = c(1,2,3,4),
                                    labels = c("Male","Female","Other","Prefer Not to Answer")))
summarytools::freq(dat$sex_sim)

gender_n <- ""
gender_percent <- ""
female_n <- nrow(filter(dat, sex_sim == "Female"))
male_n <- nrow(filter(dat, sex_sim == "Male"))
other_n <- nrow(filter(dat, sex_sim == "Other"))
pnta_n <- nrow(filter(dat, sex_sim == "Prefer Not to Answer"))
female_percent <- round(((nrow(filter(dat,sex_sim == "Female")))/(nrow(dat))*100), digits = 2)
male_percent <- round(((nrow(filter(dat, sex_sim == "Male")))/(nrow(dat))*100), digits = 2)
other_percent <- round(((nrow(filter(dat, sex_sim == "Other")))/(nrow(dat))*100), digits = 2)
pnta_percent <- round(((nrow(filter(dat, sex_sim == "Prefer Not to Answer")))/(nrow(dat))*100), digits = 2)


# recoding race factors from (1,2,3) to (black, white, Hispanic), creating a percent object for those who identify as white, and creating objects for each variable
dat <- mutate(dat, race_sim = factor(race_sim, levels = c(1,2,3),
                                     labels = c("Black","White","Hispanic")))
summarytools::freq(dat$race_sim)

race_n <- ""
race_percent <- ""
black_n <- nrow(filter(dat, race_sim == "Black"))
white_n <- nrow(filter(dat, race_sim == "White"))
hispanic_n <- nrow(filter(dat, race_sim == "Hispanic"))
black_percent <- round(((nrow(filter(dat, race_sim == "Black")))/(nrow(dat))*100), digits = 2)
white_percent <- round(((nrow(filter(dat, race_sim == "White")))/(nrow(dat))*100), digits = 2)
hispanic_percent <-round(((nrow(filter(dat, race_sim == "Hispanic")))/(nrow(dat))*100), digits = 2)
```

```{r demographics table, echo=FALSE}

# making the table
dem_table_sim <- dplyr::tribble(
  ~"Category", ~"n", ~"%",
  "Race", race_n, race_percent,
  "    Black", as.factor(black_n), as.factor(black_percent),
  "    White", as.factor(white_n), as.factor(white_percent),
  "    Hispanic", as.factor(hispanic_n), as.factor(hispanic_percent),
  "Gender", gender_n, gender_percent,
  "    Female", as.factor(female_n), as.factor(female_percent),
  "    Male", as.factor(male_n), as.factor(male_percent),
  "    Other", as.factor(other_n), as.factor(other_percent),
  "    Prefer Not to Answer", as.factor(pnta_n), as.factor(pnta_percent),
  "Grade Level", grade_n, grade_percent,
  "    Eighth Grade", as.factor(eighth_n), as.factor(eighth_percent),
  "    Tenth Grade", as.factor(tenth_n), as.factor(tenth_percent),
  "Residence", residence_n, residence_percent,
  "    Farm", as.factor(farm_n), as.factor(farm_percent),
  "    Country", as.factor(country_n), as.factor(country_percent),
  "    City", as.factor(city_n), as.factor(city_percent))

# formatting the table according to APA guidelines using the built in function
apa_table(dem_table_sim,
          caption = "Demographics", align = c("l","c","c"),
          note = "From simulated sample population. N = 14,734")

```
The sample used in this current study is a simulated data set based on the 2023 data codebook from the "Monitoring the Future" continuing study of different aspects and behaviors of high school students [@miech]. The original data set consisted of `r formatted_num` participants in either eighth or tenth grade. The simulated data set, in turn, also included `r formatted_num` simulated participants, with `r round(white_percent, digits=2)`% identifying as white, `r round(eighth_percent, digits=2)`% in eighth grade, and `r round(tenth_percent, digits=2)`% in tenth grade.

## Procedure
```{r recoding negatively keyed items, include=FALSE}

# All negatively keyed items are within the depression scale
dat <- dat %>% mutate(depression_1_sim_r = recode(depression_1_sim, `1`=3,`2`=2,`3`=1),
                      depression_3_sim_r = recode(depression_3_sim, `1`=5,`2`=4, `3`=3,`4`=2,`5`=1),
                      depression_5_sim_r = recode(depression_5_sim, `1`=5,`2`=4,`3`=3,`4`=2,`5`=1),
                      depression_7_sim_r = recode(depression_7_sim, `1`=5,`2`=4,`3`=3,`4`=2,`5`=1))
```

```{r making scales of measures, include=FALSE}

smuse_scale_sim <- dat %>% mutate(smuse_means_sim = rowMeans(pick(smuse_1_sim:smuse_4_sim)))
dat <- smuse_scale_sim

impulse_scale_sim <- dat %>% mutate(impulse_means_sim = rowMeans(pick(impulse_1_sim:impulse_14_sim)))
dat <- impulse_scale_sim

depression_scale_sim <- dat %>% mutate(depression_means_sim = rowMeans(pick(depression_1_sim_r,depression_2_sim,depression_3_sim_r,depression_4_sim,depression_5_sim_r,depression_6_sim,depression_7_sim_r,depression_8_sim,depression_9_sim,depression_10_sim)))
dat <- depression_scale_sim

anxiety_scale_sim <- dat %>% mutate(anxiety_means_sim = rowMeans(pick(anxiety_1_sim:anxiety_2_sim)))
dat <- anxiety_scale_sim
```

```{r reliability of measures, include=FALSE}

smuse_items_sim <- dat %>% dplyr::select(smuse_1_sim:smuse_4_sim)
smuse_alpha_results <- psych::alpha(smuse_items_sim)
smuse_alpha_value <- smuse_alpha_results$total$std.alpha

impulse_items_sim <- dat %>% dplyr::select(impulse_1_sim:impulse_14_sim)
impulse_alpha_results <- psych::alpha(impulse_items_sim)
impulse_alpha_value <- impulse_alpha_results$total$std.alpha

depression_items_sim <- dat %>% dplyr::select(depression_1_sim_r,depression_2_sim,depression_3_sim_r, depression_4_sim,depression_5_sim_r,depression_6_sim,depression_7_sim_r,depression_8_sim,depression_9_sim,depression_10_sim)
depression_alpha_results <- psych::alpha(depression_items_sim)
depression_alpha_value <- depression_alpha_results$total$std.alpha

anxiety_items_sim <- dat %>% dplyr::select(anxiety_1_sim,anxiety_2_sim)
anxiety_alpha_results <- psych::alpha(anxiety_items_sim)
anxiety_alpha_value <- anxiety_alpha_results$total$std.alpha

```
All demographics and measures were simulated in accordance with the codebook provided on the Monitoring the Future website [@monitori]. The codebook additionally provided the distributions for all recorded variables, which were used to determine scoring distributions for each simulated measure. For the sake of creating scales for outcome measures, a numerical scale was created, with higher values indicating higher levels of each outcome. Due to the nature of the data being entirely simulated, using Cronbach's alpha values to assess the reliability of the social media use, impulsivity, depression, and anxiety measures are not particularly informative (alpha =`r round(smuse_alpha_value, digits=4)`, `r round(impulse_alpha_value, digits=4)`, `r round(depression_alpha_value, digits=4)`, `r round(anxiety_alpha_value, digits=4)` respectively). These values are not very promising in terms of the internal reliability of the measure scales; however, with the use of the raw data, we may observe more favorable results.

## Data analysis
`r cite_r("r-references.bib")` were used for all analyses.

# Results
Given the bounds to work with raw data, I would explore how social media use is related to impulsivity, depression, and anxiety among adolescents. With the simulated data, I was able to calculate Pearson's r correlations, conduct t-tests, and create regression models to connect the simulated social media use with each behavioral and mental health outcome and how each outcome differs between eighth and tenth graders.
```{r correlations, include=FALSE}

# between social media use and impulsivity
smuse_impulse_cor_results <- cor.test(dat$smuse_means_sim, dat$impulse_means_sim)
smuse_impulse_cor_value <- smuse_impulse_cor_results$estimate
smuse_impulse_cor_pvalue <- smuse_impulse_cor_results$p.value

# between social media use and depression
smuse_depression_cor_results <- cor.test(dat$smuse_means_sim, dat$depression_means_sim)
smuse_depression_cor_value <- smuse_depression_cor_results$estimate
smuse_depression_cor_pvalue <- smuse_depression_cor_results$p.value

# between social media use and anxiety
smuse_anx_cor_results <- cor.test(dat$smuse_means_sim, dat$anxiety_means_sim)
smuse_anx_cor_value <- smuse_anx_cor_results$estimate
smuse_anx_cor_pvalue <- smuse_anx_cor_results$p.value

```

For the sake of finding significance in the correlations between measures, 0.05 was used as a threshold for the p-values. Simulated social media use and impulse control were found to be weakly negatively correlated with a p-value above the threshold, indicating that either the nature of the simulated data are not representative of authentic data or the hypothesis that social media use and impulsivity are positively correlated can likely be disproved (Pearson's r = `r round(smuse_impulse_cor_value, digits = 3)`, *p* = `r printp(smuse_impulse_cor_pvalue)`). Social media use and depression values were produced a correlation value of practically zero with a p-value above the threshold, so it is a near negligible relationship based on the simulated data (Pearson's r = `r sprintf("%.3f", smuse_depression_cor_value)`, *p* = `r printp(smuse_depression_cor_pvalue)`). Lastly, it was found that social media use and anxiety are weakly negatively correlated based on the simulated data, also indicating a weak relationship that can disprove the hypothesis formed for social media use and anxiety or an inaccurate representation of how the measures can be seen using raw data and participants (Pearson's r = `r smuse_anx_cor_value`, *p* = `r printp(smuse_anx_cor_pvalue)`).

*Figure 1.*


```{r box plot, echo=FALSE, warning=FALSE, fig.align='left', fig.cap = '', results='asis'}

plot_colors <- c("#F8766D","#7CAE00","#00BFC4","#C77CFF")

# pivoting the data from wide to long format
data_long <- dat %>% pivot_longer(
  cols = c(anxiety_means_sim, depression_means_sim, impulse_means_sim, smuse_means_sim),
  names_to = "Scale",
  values_to = "Value"
)

# making the boxplot
full_plot <- ggplot(data_long, aes(x = Scale, y = Value), na.rm = TRUE)+
  geom_boxplot(fill = plot_colors)+
  theme_apa()+
  labs(title = "Outcome Scale Values")+
  scale_x_discrete(labels = c("anxiety_means_sim" = "Anxiety", "depression_means_sim" = "Depression", "impulse_means_sim" = "Impulsivity", "smuse_means_sim" = "Social Media Use"))
full_plot
```

```{r t tests, include=FALSE}

# these will be comparing eighth (n1) and tenth graders (n2) on measures
t_test_df_sim <- ((nrow(filter(dat, grade_sim == "Eighth Grade")))+(nrow(filter(dat, grade_sim == "Tenth Grade")))-2)
critical_t_value <- qt(.025, t_test_df_sim, lower.tail=TRUE)*2
critical_t_value

# social media use t test with objects made of the t stat, p-value, and group means for text insertion
smuse_t_test <- t.test(dat$smuse_means_sim~dat$grade_sim)
smuse_t_value <- smuse_t_test$statistic
smuse_t_pvalue <- smuse_t_test$p.value
smuse_t_mean_8 <- mean(dat$smuse_means_sim, grade_sim = "Eighth Grade", na.rm = TRUE)
smuse_t_sd_8 <- sd(dat$smuse_means_sim[dat$grade_sim == "Eighth Grade"], na.rm = TRUE)
smuse_t_mean_10 <- mean(dat$smuse_means_sim, grade_sim = "Tenth Grade", na.rm = TRUE)
smuse_t_sd_10 <- sd(dat$smuse_means_sim[dat$grade_sim == "Tenth Grade"], na.rm = TRUE)

# impulsivity t test with objects made of the t stat, p-value, and group means for text insertion
impulse_t_test <- t.test(dat$impulse_means_sim~dat$grade_sim)
impulse_t_value <- impulse_t_test$statistic
impulse_t_pvalue <- impulse_t_test$p.value
impulse_t_mean_8 <- mean(dat$impulse_means_sim, grade_sim = "Eighth Grade", na.rm = TRUE)
impulse_t_sd_8 <- sd(dat$impulse_means_sim[dat$grade_sim == "Eighth Grade"], na.rm = TRUE)
impulse_t_mean_10 <- mean(dat$impulse_means_sim, grade_sim = "Tenth Grade", na.rm = TRUE)
impulse_t_sd_10 <- sd(dat$impulse_means_sim[dat$grade_sim == "Tenth Grade"], na.rm = TRUE)

# depression t test with objects made of the t stat, p-value, and group means for text insertion
depression_t_test <- t.test(dat$depression_means_sim~dat$grade_sim)
depression_t_value <- depression_t_test$statistic
depression_t_pvalue <- depression_t_test$p.value
depression_t_mean_8 <- mean(dat$depression_means_sim, grade_sim = "Eighth Grade", na.rm = TRUE)
depression_t_sd_8<- sd(dat$depression_means_sim[dat$grade_sim == "Eighth Grade"], na.rm = TRUE)
depression_t_mean_10 <- mean(dat$depression_means_sim, grade_sim = "Tenth Grade", na.rm = TRUE)
depression_t_sd_10 <- sd(dat$depression_means_sim[dat$grade_sim == "Tenth Grade"], na.rm = TRUE)

# anxiety t test with objects made of the t stat, p-value, and group means for text insertion
anxiety_t_test <- t.test(dat$anxiety_means_sim~dat$grade_sim)
anxiety_t_value <- anxiety_t_test$statistic
anxiety_t_pvalue <- anxiety_t_test$p.value
anxiety_t_mean_8 <- mean(dat$anxiety_means_sim, grade_sim = "Eighth Grade", na.rm = TRUE)
anxiety_t_sd_8 <- sd(dat$anxiety_means_sim[dat$grade_sim == "Eighth Grade"], na.rm = TRUE)
anxiety_t_mean_10 <- mean(dat$anxiety_means_sim, grade_sim = "Tenth Grade", na.rm = TRUE)
anxiety_t_sd_10 <- sd(dat$anxiety_means_sim[dat$grade_sim == "Tenth Grade"], na.rm = TRUE)
```

T-tests were performed to compare how participants scored on social media use, impulsivity, depression, and anxiety scales based on grade level. Most items were scored on a scale from 1 to 5, with 1 being lower levels of the measure and 5 being high levels of the measure. For social media use, there was not a particularly large observed difference between how eighth and tenth graders were scoring, with the average group means at `r smuse_t_mean_8` with a standard deviation of `r smuse_t_sd_8` for eighth grade and `r smuse_t_mean_10` with a standard deviation of `r smuse_t_sd_10` for tenth grade (t-statistic = `r smuse_t_value`, *p* = `r printp(smuse_t_pvalue)`). Similar results were found for impulsivity, and while the difference between the grade levels was found to be higher, the average scores were also quite similar. For eighth grade the average score was `r impulse_t_mean_8` with a standard deviation of `r impulse_t_sd_8` and for tenth grade the average score was `r impulse_t_mean_10` with a standard deviation of `r impulse_t_sd_10`. The p-value for this comparison denoted a lack of significance (t-statistic = `r impulse_t_value`, *p* = `r printp(impulse_t_pvalue)`). The depression measure also saw comparable results, with average scores of `r depression_t_mean_8` with a standard deviation of `r depression_t_sd_8` for eighth grade and `r depression_t_mean_10` with a standard deviation of `r depression_t_sd_10` for tenth grade (t-statistic = `r depression_t_value`, *p* = `r printp(depression_t_pvalue)`). Comparing the grade levels on the anxiety measure again saw similar group means for the measure separated by grade, with eighth and tenth graders averaging scores of `r anxiety_t_mean_8` and `r anxiety_t_mean_10` with standard deviations of `r anxiety_t_sd_8` and `r anxiety_t_sd_10`, respectively (t-statistic = `r anxiety_t_value`, *p* = `r printp(anxiety_t_pvalue)`). These results showed very little difference between how eighth and tenth graders scored on any of the simulated behavioral or mental health measures. Overall, from analyzing simulated data of the observed measures, no significant relationships were found between youth social media use and impulsivity, depression, or anxiety, and no significant differences were found between how eighth and tenth grade participants scored on the measures.

# Discussion
The purpose of this study was to mimic the workflow involved in creating reproducible research projects. All analyses were done using simulated data, so results are not representative of the true relationships between these variables that are observed in the raw data from 2023. Through the simulated analyses, no significant relationships between social media use and impulsivity, depression, or anxiety were noted, nor was a difference found between how eighth and tenth tenth-grade participants scored on any of the observed measures. Therefore, none of the study hypotheses were supported by the simulated data, and the raw data should be used to examine these relationships in a future analysis.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
