---
title             : "Social Media Use in Youth and Related Mental Health and Behavioral Outcomes"
shorttitle        : "Youth Social Media Use"

author: 
  - name          : "Natalie Cunningham"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "75 E River Pkwy, Minneapolis, MN 55455"
    email         : "cunni734@umn.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "University of Minnesota"
  - id            : "2"
    institution   : "Konstanz Business School"

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  One or two sentences to put the results into a more **general context**.
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library(papaja)
library(groundhog)
library(dplyr)
library(psych)
library(labelled)
library(ggplot2)
library(apaTables)
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(1980)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

```{r data setup, include=FALSE}
dat <- read.csv("../Data/2025-11-13_youth-media-use_simulated-data.csv")
names(dat)
```

# Introduction

Technology is rapidly advancing and becoming more accessible to the adolescents of ours and future generations. With this accessibility, the doors to social media access and use are wide open to youth. Additionally, mental health and behavioral issues among youth are being more widely addressed, and many claims have been made that social media, and media use in general, have negative effects on children's well-being. These technologies and media platforms will continue to evolve, and, therefore, it is becoming increasingly necessary to understand the psychological impacts of these growing entities on youth. The focus of this study is to use reproducible workflows to examine the relationship between social media use and mental health and behavioral outcomes, specifically impulsivity, depression, and anxiety, among a population of high-school aged youth. The three hypotheses tested are as follows:

-   Social media use will be positively correlated with impulsivity, and higher use of social media will lead to higher levels of impulsivity.

-   Social media use will be positively correlated with feelings of depression, and higher use of social media will lead to higher levels of depression.

-   Social media use will be positively correlated with feelings of anxiety, and higher use of social media will lead to higher levels of anxiety.

To asses these, analyses were done on a simulated data set compiled referencing the data codebook from the 2023 Monitoring the Future Main Study, conducted on eighth and tenth grade participants by the University of Michigan. This study analyzed the relationships between social media use, impulsivity, depression, and anxiety using this simulated data to outline the procedure and method by which these research questions could be assessed in a reproducible manner.

# Methods

The current study was not preregistered and raw data was not collected or analyzed. Simulated data was created based on raw data from the 2023 study on eighth and tenth graders from the “Monitoring the Future” continuing study. Simulated data and code are available at [this hyperlink to the github repository](https://github.com/cunningpsych/Social-Media-Use_in_Youth).

## Participants

```{r demographics, include=FALSE}
# recoding grade factors from (1,2) to (eighth grade, tenth grade) and creating percent objects for each grade
dat <- mutate(dat, grade_sim = factor(grade_sim, levels = c(1,2),
                                      labels = c("Eighth Grade", "Tenth Grade")))

summarytools::freq(dat$grade_sim)
eighth_percent <- (nrow(filter(dat, grade_sim == "Eighth Grade"))/(nrow(dat))*100)
tenth_percent <- (nrow(filter(dat, grade_sim == "Tenth Grade"))/(nrow(dat))*100)

# recoding residence factors from (1,2,3) to (farm, country, citytown)
dat <- mutate(dat, residence_sim = factor(residence_sim, levels = c(1,2,3),
                                          labels = c("Farm", "Country", "City")))
summarytools::freq(dat$residence_sim)


# recoding sex factors from (1,2,3,4) to (male, female, other, prefer not to answer)
dat <- mutate(dat, sex_sim = factor(sex_sim, levels = c(1,2,3,4),
                                    labels = c("Male","Female","Other","Prefer Not to Answer")))
summarytools::freq(dat$sex_sim)

# recoding race factors from (1,2,3) to (black, white, hispanic), creating a percent object for those who identify as white, and creating objects for each variable
dat <- mutate(dat, race_sim = factor(race_sim, levels = c(1,2,3),
                                     labels = c("Black","White","Hispanic")))
summarytools::freq(dat$race_sim)

black_percent <- (nrow(filter(dat, race_sim == "Black"))/(nrow(dat))*100)
white_percent <- (nrow(filter(dat, race_sim == "White"))/(nrow(dat))*100)
hispanic_percent <- (nrow(filter(dat, race_sim == "Hispanic"))/(nrow(dat))*100)

black_pop <- nrow(filter(dat, race_sim == "Black"))
white_pop <- nrow(filter(dat, race_sim == "White"))
hispanic_pop <- nrow(filter(dat, race_sim == "Hispanic"))
```

The sample used in this current study is a simulated data set based on the 2023 data codebook from the "Monitoring the Future" continuing study of different aspects and behaviors of high school students. The original data set consisted of `r nrow(dat)` participants in either eighth or tenth grade. The simulated data set, in turn, also included `r nrow(dat)` simulated participants, with `r round(white_percent, digits=2)`% identifying as white, `r round(eighth_percent, digits=2)`% in eighth grade, and `r round(tenth_percent, digits=2)`% in tenth grade.

```{r demographics table, include=FALSE}
# making the table
race_table_sim <- dplyr::tribble(
  ~"Racial Group", ~"n", ~"Percentage",
  "Black", black_pop, black_percent,
  "White", white_pop, white_percent,
  "Hispanic", hispanic_pop, hispanic_percent)

# making the table a dat frame
data.frame(race_table_sim)


# formatting the table according to APA guidelines
apa_table(race_table_sim,
          caption = "Demographics", align = c("l","c","c"),
          placement = "h",
          note = "From simulated sample population")
```

## Procedure

```{r recoding negatively keyed items, include=FALSE}

# All negatively keyed items are within the depression scale
dat <- dat %>% mutate(depression_1_sim_r = recode(depression_1_sim, `1`=3,`2`=2,`3`=1),
                      depression_3_sim_r = recode(depression_3_sim, `1`=5,`2`=4, `3`=3,`4`=2,`5`=1),
                      depression_5_sim_r = recode(depression_5_sim, `1`=5,`2`=4,`3`=3,`4`=2,`5`=1),
                      depression_7_sim_r = recode(depression_7_sim, `1`=5,`2`=4,`3`=3,`4`=2,`5`=1))
```

```{r making scales of measures, include=FALSE}

smuse_scale_sim <- dat %>% mutate(smuse_means_sim = rowMeans(pick(smuse_1_sim:smuse_4_sim)))
dat <- smuse_scale_sim

impulse_scale_sim <- dat %>% mutate(impulse_means_sim = rowMeans(pick(impulse_1_sim:impulse_14_sim)))
dat <- impulse_scale_sim

depression_scale_sim <- dat %>% mutate(depression_means_sim = rowMeans(pick(depression_1_sim_r,depression_2_sim,depression_3_sim_r,depression_4_sim,depression_5_sim_r,depression_6_sim,depression_7_sim_r,depression_8_sim,depression_9_sim,depression_10_sim)))
dat <- depression_scale_sim

anxiety_scale_sim <- dat %>% mutate(anxiety_means_sim = rowMeans(pick(anxiety_1_sim:anxiety_2_sim)))
dat <- anxiety_scale_sim
```

```{r reliability of measures, include=FALSE}

smuse_items_sim <- dat %>% dplyr::select(smuse_1_sim:smuse_4_sim)
smuse_alpha_results <- psych::alpha(smuse_items_sim)
smuse_alpha_value <- smuse_alpha_results$total$std.alpha

impulse_items_sim <- dat %>% dplyr::select(impulse_1_sim:impulse_14_sim)
impulse_alpha_results <- psych::alpha(impulse_items_sim)
impulse_alpha_value <- impulse_alpha_results$total$std.alpha

depression_items_sim <- dat %>% dplyr::select(depression_1_sim_r,depression_2_sim,depression_3_sim_r, depression_4_sim,depression_5_sim_r,depression_6_sim,depression_7_sim_r,depression_8_sim,depression_9_sim,depression_10_sim)
depression_alpha_results <- psych::alpha(depression_items_sim)
depression_alpha_value <- depression_alpha_results$total$std.alpha

anxiety_items_sim <- dat %>% dplyr::select(anxiety_1_sim,anxiety_2_sim)
anxiety_alpha_results <- psych::alpha(anxiety_items_sim)
anxiety_alpha_value <- anxiety_alpha_results$total$std.alpha

```

Due to the nature of the data being entirely simulated, using Cronbach's alpha values to assess the reliability of the social media use, impulsivity, depression, and anxiety measures are not particularly informative (alpha =`r round(smuse_alpha_value, digits=4)`, `r round(impulse_alpha_value, digits=4)`, `r round(depression_alpha_value, digits=4)`, `r round(anxiety_alpha_value, digits=4)` respectively). These values are not very promising in terms of the internal reliability of the measure scales, however, with the use of genuine raw data, we may observe more favorable results.

# Results

Given the bounds to work with raw data, I would plan to explore how social media use is related to impulsivity, depression, and anxiety among adolescents. With the simulated data, I was able to calculate Pearson's r correlations, conduct t-tests, and create regression models to connect the simulated social media use with each behavioral and mental health outcome and how each outcome differs between eighth and tenth graders.

```{r correlations, include=FALSE}

# between social media use and impulsivity
smuse_impulse_cor_results <- cor.test(dat$smuse_means_sim, dat$impulse_means_sim)
smuse_impulse_cor_value <- smuse_impulse_cor_results$estimate
smuse_impulse_cor_pvalue <- smuse_impulse_cor_results$p.value

# between social media use and depression
smuse_depression_cor_results <- cor.test(dat$smuse_means_sim, dat$depression_means_sim)
smuse_depression_cor_value <- smuse_depression_cor_results$estimate
smuse_depression_cor_pvalue <- smuse_depression_cor_results$p.value

# between social media use and anxiety
smuse_anx_cor_results <- cor.test(dat$smuse_means_sim, dat$anxiety_means_sim)
smuse_anx_cor_value <- smuse_anx_cor_results$estimate
smuse_anx_cor_pvalue <- smuse_anx_cor_results$p.value

```

For the sake of finding significance in the correlations between measures, 0.05 was use as a threshold for the p-values. Simulated social media use and impulse control were found to be weakly negatively correlated with a p-value above the threshold, indicating that either the nature of the simulated data is not representative of authentic data or the hypothesis that social media use and impulsivity are positively correlated can likely be disproved (Pearson's r = `r round(smuse_impulse_cor_value, digits = 3)`, p = `r round(smuse_impulse_cor_pvalue, digits = 3)`). Social media use and depression values were weakly positively correlated with a p-value above the threshold, indicating that while the correlation is positive, as predicted, it is not a strong relationship (Pearson's r = `r round(smuse_depression_cor_value, digits = 3)`, p = `r round(smuse_depression_cor_pvalue, digits=3)`). Lastly, it was found that social media use and anxiety are _______ correlated based on the simulated data,

```{r t tests, include=FALSE}

# these will be comparing eighth (n1) and tenth graders (n2) on measures
t_test_df_sim <- ((nrow(filter(dat, grade_sim == "Eighth Grade")))+(nrow(filter(dat, grade_sim == "Tenth Grade")))-2)
critical_t_value <- qt(.025, t_test_df_sim, lower.tail=TRUE)*2
critical_t_value

# social media use t test with objects made of the t stat, p-value, and group means for text insertion
smuse_t_test <- t.test(dat$smuse_means_sim~dat$grade_sim)
smuse_t_value <- smuse_t_test$statistic
smuse_t_pvalue <- smuse_t_test$p.value
smuse_t_means <- smuse_t_test$estimate

# impulsivity t test with objects made of the t stat, p-value, and group means for text insertion
impulse_t_test <- t.test(dat$impulse_means_sim~dat$grade_sim)
impulse_t_value <- impulse_t_test$statistic
impulse_t_pvalue <- impulse_t_test$p.value
impulse_t_means <- impulse_t_test$estimate

# depression t test with objects made of the t stat, p-value, and group means for text insertion
depression_t_test <- t.test(dat$depression_means_sim~dat$grade_sim)
depression_t_value <- depression_t_test$statistic
depression_t_pvalue <- depression_t_test$p.value
depression_t_means <- depression_t_test$estimate

# anxiety t test with objects made of the t stat, p-value, and group means for text insertion
anxiety_t_test <- t.test(dat$anxiety_means_sim~dat$grade_sim)
anxiety_t_value <- anxiety_t_test$statistic
anxiety_t_pvalue <- anxiety_t_test$p.value
anxiety_t_means <- anxiety_t_test$estimate

```

```{r regression analyses, include=FALSE}
# the regression models will be to assess whether social media use could predict the mental health and behavioral outcomes that were measured

# social media use on impulsivity
smuse_impulse_reg_model <- summary(lm(impulse_means_sim~smuse_means_sim, data = dat))
smuse_impulse_reg_pvalue <- smuse_impulse_reg_model$coefficients[, 4]

# social media use on depression
smuse_depression_reg_model <- summary(lm(depression_means_sim~smuse_means_sim, data = dat))
smuse_depression_reg_pvalue <- smuse_depression_reg_model$coefficients[,4]

# social media use on anxiety
smuse_anx_reg_model <- summary(lm(anxiety_means_sim~smuse_means_sim, data = dat))
smuse_anx_reg_pvalue <- smuse_anx_reg_model$coefficients[,4]

```

```{r visualizations, echo=FALSE, fig.pos= "ht", out.width= "\\textwidth"}
ggplot(dat, aes(x=smuse_means_sim, y=impulse_means_sim))+
  geom_point(alpha=.25, na.rm = TRUE)+
  geom_smooth(formula = y~x, method = "lm", na.rm = TRUE)+
  theme_apa()+
  labs(title= "Relationship between Social Media Use and Impulsivity", x="Social Media Use", y="Impulsivity")

ggplot(dat, aes(x=depression_means_sim, y=impulse_means_sim))+
  geom_point(alpha=.25, na.rm=TRUE)+
  geom_smooth(formula = y~x, method = "lm", na.rm = TRUE)+
  theme_apa()+
  labs(title= "Relationship between Impulsivity and Depression", x="Depression", y="Impulsivity")


```

# Discussion

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
